---
title: 'Homework 2: Multiple Correspondence Analysis and Clustering'
author: 'Group 13 : Anna Monsó Rodríguez, Walter J. Troiani Vargas, Joan Acero Pousa'
date: "2024-11-24"
output:
  pdf_document: default
  html_document: default
---

# Exploratory Data Analysis

## A) Import the data set correctly to R and assign type of each variable correctly 
```{r}

```

## B) Convert te numerical variables to factors consist of two categories "low" and "high" by cutting them from their median and save them as new variables to the data set

```{r}

```



# Multiple Correspondence Analysis

## C) Apply MCA on the categorical variables taking Region and Chanel as supplementary variables by using MCA() function in FactoMineR package. Interpret your findings 
```{r}

```

# Hierarchical Clustering

## D) Apply hierarchical clustering on MCA scores obtained from previous step by using HCPC function. How many clusters are constructed?

```{r}

```


## E) Apply hierarchical clustering on PCA scores considering only the nhumerical variables and taking Region and Channel variables as supplementary variables. How many clusters are constructed.

```{r}

```
# Profiling Analysis

## F) Interpret the clusters obtained from both methods



## G) Whcih of the above hierarchical clustering methods would you choose? Why?

